---
title: "Blood analysis"
output: html_notebook
---

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r Dane}
library(dplyr)
library(ggplot2)
library(tidyr)

#cardio <- read.csv("cardio_train.csv", sep = ";")
#cardio_cleaned <- read.csv("cardio_cleaned.csv")



```

```{r}

zmienne <- c("height", "weight", "ap_hi", "ap_lo", "cholesterol")
for(zmienna in zmienne){
  pom <- cardio %>%  
  filter(between(.data[[zmienna]], quantile(.data[[zmienna]], probs = 0.05), quantile(.data[[zmienna]], probs = 0.95))) %>% 
  mutate(!!sym(zmienna) := .data[[zmienna]] + runif(length(.data[[zmienna]]), min = 0, max = 0.1*max(.data[[zmienna]])), 
         age = age + runif(length(age), min = 0, max = 0.1*max(age)))
  
  p <- ggplot(pom, aes(x = age, y = .data[[zmienna]])) +
          geom_point(shape = 16, color = rgb(pom$cardio, 0, 0, 0.5)) +
          labs(title = "testing", x = "age", y = zmienna) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
        print(p)
  print(cor(pom[,zmienna], pom$age))
}

pom2 <- cardio %>% group_by(gender) %>% summarise(count=n())

barplot(pom2$count, pom2$gender, names.arg = pom2$gender)


pom
ggplot()+
  geom_boxplot()

```

```{r}


zmienne <- c("height", "weight", "ap_hi", "ap_lo")
for(zmienna in zmienne){
  pom <- cardio_cleaned[is.na(cardio_cleaned)] <- 0
  pom <- cardio_cleaned %>%
  filter(between(.data[[zmienna]], quantile(.data[[zmienna]], probs = 0.05), quantile(.data[[zmienna]], probs = 0.95)) ) %>% 
  mutate(!!sym(zmienna) := .data[[zmienna]] + runif(length(.data[[zmienna]]), min = 0, max = 0.01*max(.data[[zmienna]])), 
         age = age + runif(length(age), min = 0, max = 0.01*max(age)))
  
  p <- ggplot(pom, aes(x = age, y = .data[[zmienna]])) +
          geom_point(shape = 16, color = rgb(pom$cardio, 0, 0, 0.5)) +
          labs(title = "testing", x = "age", y = zmienna) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
        print(p)
  print(cor(pom[,zmienna], pom$age))
}
```


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)



usa_death_data <- read.csv("slctmort.csv", sep = ";")

colnames(usa_death_data) <- substr(colnames(usa_death_data), 2, 5)
colnames(usa_death_data)[1] <- "Races"
#View(usa_death_data)

usa_death_data_long <- usa_death_data %>%
  filter(grepl("All causes|Diseases of heart", Races)) %>% 
  pivot_longer(cols = colnames(usa_death_data)[-1],
               names_to = "Year",
               values_to = "Deaths") %>% 
  mutate_all(~ replace(., . == "" | . == "- - -" | . == "…", NA)) %>% 
  mutate(Deaths = as.integer(gsub(",", "", Deaths)),
         Year = as.numeric(Year))


```

```{r}



usa_death_data_long %>%
      ggplot(aes(x = as.factor(Year), y = Deaths, color = Races)) +
      geom_point(size = 1) +
      geom_line() +  # Optionally add lines to connect points
      labs(title = "Death Counts by Year ",
           x = "Year", 
           y = "Number of Deaths",
           color = "Cause of Death") +
      theme(legend.position="none")

Races_causes <- unique(usa_death_data_long$Races)

usa_death_data_long %>% 
  #filter(Year > 1990) %>% 
  filter(Races %in% Races_causes[c(1, 2)]) %>% 
      ggplot(aes(x = Year, y = Deaths, color = as.factor(Races))) +
      geom_point(size = 1) +
      geom_line() +  # Optionally add lines to connect points
      ylim(c(0, max(usa_death_data_long$Deaths)))+
      labs(title = "Death Counts by Year ",
           x = "Year", 
           y = "Number of Deaths",
           color = "Cause of Death")

usa_death_data_long %>% 
  filter(Year >= 1990) %>% 
  filter(Races %in% Races_causes[c(7, 8, 9, 10, 11, 12, 27, 28)]) %>% 
      ggplot(aes(x = Year, y = Deaths, color = as.factor(Races))) +
      geom_point(size = 1) +
      geom_line() +  # Optionally add lines to connect points
      ylim(c(0, max(usa_death_data_long$Deaths)))+
      labs(title = "Death Counts by Year ",
           x = "Year", 
           y = "Number of Deaths",
           color = "Cause of Death")


usa_death_data_long %>% 
  filter(Year >= 1990) %>% 
  filter(Races %in% Races_causes[c(7, 8, 11, 12)]) %>% 
      ggplot(aes(x = Year, y = Deaths, color = as.factor(Races))) +
      geom_point(size = 1) +
      geom_line() +  # Optionally add lines to connect points
      ylim(c(0, max(usa_death_data_long$Deaths)))+
      labs(title = "Death Counts by Year ",
           x = "Year", 
           y = "Number of Deaths",
           color = "Cause of Death")
```

```{r}
Races_causes <- unique(usa_death_data_long$Races)
pom<-usa_death_data_long %>% 
  filter(Year >= 1990) %>% 
  filter(Races %in% Races_causes[c(7, 8, 11, 12)]) %>% 
  mutate(Races = if_else(Races == "Black or African American All causes", "Black_AllCauses", 
                 if_else(Races == "Black or African American Diseases of heart", "Black_HeartDiseases",
                 if_else(Races == "Race, Hispanic origin, and cause of death White All causes", "White_AllCauses", 
                 if_else(Races == "Race, Hispanic origin, and cause of death White Diseases of heart", "White_HeartDiseases", "")))))

pom %>% 
  ggplot(aes(x = Year, y = Deaths, color = as.factor(Races))) +
  geom_point(size = 1) +
  geom_line() +  # Optionally add lines to connect points
  ylim(c(0, max(usa_death_data_long$Deaths)))+
  labs(title = "Death Counts by Year ",
       x = "Year", 
       y = "Number of Deaths",
       color = "Cause of Death")


pom %>% 
  pivot_wider(names_from = Races, values_from = Deaths) %>% 
  mutate(White_DeathsRatio = White_HeartDiseases/White_AllCauses,
         Black_DeathsRatio = Black_HeartDiseases/Black_AllCauses) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = White_DeathsRatio, color = "White"), linewidth = 1) +
  geom_line(aes(y = Black_DeathsRatio, color = "Black"), linewidth = 1) +
  ylim(0, NA) +
  labs(title = "Death ratios for White and Black",
       x = "Year",
       y = "Ratio",
       color = "Race") +  # Label for legend
  scale_color_manual(values = c("White" = "#ffffff", "Black" = "#721f14")) +  # Manual color specification
  guides(color = guide_legend(
    override.aes = list(
      color = c("#721f14", "white")  # Changes "White" legend text color to gray, "Black" remains black
    )
  )) +
  theme(
    text = element_text(size = 25),
    axis.text.x = element_text(angle = 0, hjust = 1),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    plot.title = element_blank(),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent"),
    legend.box.background = element_rect(fill = "transparent", color = NA),
    legend.key = element_rect(fill = "transparent")
  ) -> p
print(p)


  ggsave(
  plot = p,
  filename = "Races-plot.png",
  bg = "transparent"
)
 
  
  pom %>% 
  pivot_wider(names_from = Races, values_from = Deaths) %>% 
  mutate(White_DeathsRatio = White_HeartDiseases/White_AllCauses,
         Black_DeathsRatio = Black_HeartDiseases/Black_AllCauses) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = White_DeathsRatio, color = "White"), linewidth = 1) +
  geom_line(aes(y = Black_DeathsRatio, color = "Black"), linewidth = 1) +
  ylim(0, NA) +
  labs(title = "Death ratios for White and Black",
       x = "Year",
       y = "Ratio",
       color = "Race") +  # Label for legend
  scale_color_manual(values = c("White" = "#ffffff", "Black" = "#721f14")) +  # Manual color specification
  guides(color = guide_legend(
    override.aes = list(
      color = c("#721f14", "white")  # Changes "White" legend text color to gray, "Black" remains black
    )
  )) +
  theme(
    text = element_text(size = 25, color = "white", family = "Playfair Display"),  # Set color explicitly to white and font family
    axis.text.x = element_text(angle = 0, hjust = 1, color = "white"),  # Explicitly set axis text color
    axis.text = element_text(color = "white"),  # Ensure axis text is white
    axis.title = element_text(color = "white"),  # Ensure axis titles are white
    plot.title = element_blank(),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    panel.background = element_rect(fill = "#a09386", colour = NA),
    plot.background = element_rect(fill = "#a09386", colour = NA),
    legend.background = element_rect(fill = "#a09386"),
    legend.box.background = element_rect(fill = "#a09386", color = NA),
    legend.key = element_rect(fill = "#a09386"),
    plot.margin = margin(0, 0, 0, 0)  # Optional: remove margins to prevent transparent borders
  ) -> p2
  
  
  print(p2)
    ggsave(
  plot = p2,
  filename = "Races-plot_v2.png"#,bg = "transparent"
)

  

```



```{r}



for(i in 1:((length(colnames(usa_death_data))-1)/2) ){
  usa_death_data_long <- usa_death_data %>%
    pivot_longer(cols = c(.data[[colnames(usa_death_data)[[i*2-1]]]], .data[[colnames(usa_death_data)[[i*2]]]]),
                 names_to = "Cause",
                 values_to = "Deaths")
  
  
  # Plot both variables on the same plot with different colors
  p<-usa_death_data_long %>%
      ggplot(aes(x = Year, y = Deaths, color = Cause)) +
      geom_point(size = 1) +
      geom_line() +  # Optionally add lines to connect points
      labs(title = paste("Death Counts by Year ", colnames(usa_death_data)[[i*2-1]]),
           x = "Year", 
           y = "Number of Deaths",
           color = "Cause of Death") +
      theme_minimal()
  print(p)
}

```

```{r}

library(dplyr)
library(ggplot2)
library(tidyr)

frmgham <- read.csv("frmgham2.csv")
str(frmgham)

frmgham <- frmgham %>% mutate(SEX = if_else(SEX==1, "Male", "Female"))


```


```{r}


zmienne <- colnames(frmgham)[c(3,4,5, 8, 9, 12, 13, 20, 22, 23)]
for(j in zmienne){
  for(i in zmienne){
    frmgham[is.na(frmgham)] <- 0
    #pom <- frmgham %>%
    #filter(between(.data[[zmienna]], quantile(.data[[zmienna]], probs = 0.05), quantile(.data[[zmienna]], probs = 0.95)) ) %>% 
    #mutate(!!sym(zmienna) := .data[[zmienna]] + runif(length(.data[[zmienna]]), min = 0, max = 0.01*max(.data[[zmienna]])), 
    #       age = age + runif(length(age), min = 0, max = 0.01*max(age)))
    
    p <- ggplot(frmgham, aes(x = .data[[i]], y = .data[[j]])) +
            geom_point(shape = 16, color = rgb(frmgham$CVD, 0, 0, 0.5)) +
            labs(title = cor(frmgham[,i], frmgham[,j]) , x = i, y = j) +
            theme(axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1))    
    print(p)
    print(paste(i,"  ",j,"  ",cor(frmgham[,i], frmgham[,j]), 
                ifelse(abs(cor(frmgham[,i], frmgham[,j]))>0.3, "AAAAAAAAAAAAAAAa", "")))
  }
}

```

```{r}

normalise <- function(x, na.rm = FALSE) ((x-min(x, na.rm = na.rm)) / 
                          (max(x, na.rm = na.rm)-min(x, na.rm = na.rm)))

frmgham_long <- frmgham %>% 
  mutate_at(colnames(frmgham)[3:13], ~normalise(., na.rm = TRUE)) %>% 
  pivot_longer(cols = colnames(frmgham)[c(29, 27, 10, 8, 11)], 
               names_to = "Variable", 
               values_to = "Value")

frmgham_long %>% 
  ggplot(aes(x=Variable, y=Value, colour = as.factor(SEX)))+
  geom_boxplot()+
  labs(title = "title", x= "vars",y= "normalised")+
  theme(axis.text.x = element_text(angle = 60, hjust=1),
        axis.title.y = element_text(angle = 0))

frmgham_long <- frmgham %>% 
  #mutate(across(colnames(frmgham)[3:13], ~ normalise(.x, na.rm = TRUE))) %>% 
  
  pivot_longer(cols = colnames(frmgham)[3:13], 
               names_to = "Variable", 
               values_to = "Value")

frmgham %>% 
  filter(CIGPDAY==0) %>% 
  #group_by(RANDID, SEX, CIGPDAY) %>% 
  ggplot(aes(x=SEX))+
  geom_bar()

frmgham %>% 
  filter(CIGPDAY>0) %>% 
  #group_by(RANDID, SEX, CIGPDAY) %>% 
  ggplot(aes(x = SEX, y=CIGPDAY))+
  geom_boxplot()


frmgham %>% 
  filter(CIGPDAY==0) %>% 
  #group_by(RANDID, SEX, CIGPDAY) %>% 
  ggplot(aes(x=as.factor(DEATH)))+
  geom_bar()

frmgham %>% 
  filter(CIGPDAY>0) %>% 
  #group_by(RANDID, SEX, CIGPDAY) %>% 
  ggplot(aes(x = as.factor(DEATH), y=CIGPDAY))+
  geom_boxplot()



frmgham %>% 
  filter(CIGPDAY==0) %>% 
  #group_by(RANDID, SEX, CIGPDAY) %>% 
  ggplot(aes(x=as.factor(PREVCHD)))+
  geom_bar()

frmgham %>% 
  filter(CIGPDAY>0) %>% 
  #group_by(RANDID, SEX, CIGPDAY) %>% 
  ggplot(aes(x = as.factor(PREVCHD), y=CIGPDAY))+
  geom_boxplot()


#frmgham %>% 
#  filter(CIGPDAY>0) %>% 
#  ggplot(aes(x=CIGPDAY, fill = SEX))+
#  geom_density(alpha = 0.7)



```

```{r}
#frmgham <- frmgham %>% filter(complete.cases(.))


frmgham %>%
  ggplot(aes(y=AGE, x=as.character(SEX))) +
  geom_boxplot(alpha=0.5)


frmgham %>%
  ggplot(aes(x=AGE)) +
  geom_density()

frmgham %>%
  ggplot(aes(x=AGE, fill=as.character(CIGPDAY %/% 25))) +
  geom_density(alpha=0.5)

```


```{r}
frmgham %>%  
  group_by(SEX) %>% 
  summarise("Stroke yes-no" = mean(STROKE),
            mean(MI_FCHD),
            mean(DIABETES),
            mean(CIGPDAY, na.rm = TRUE),
            "Use of Anti-hypertensive medication" = mean(BPMEDS, na.rm = TRUE)) -> pom  

# Compute proportional difference between Female and Male values for each variable

proportional_diff <- (pom[pom$SEX == "Female", -1] - pom[pom$SEX == "Male", -1]) / 
  (pom[pom$SEX == "Female", -1] + pom[pom$SEX == "Male", -1])

# Convert to a data frame and set column names
proportional_diff <- as.data.frame(proportional_diff)
colnames(proportional_diff) <- colnames(pom)[-1]

# Reshape the data into a long format for ggplot
proportional_diff_long <- proportional_diff %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value")

# Plot the proportional difference with color indicating sex
ggplot(proportional_diff_long, aes(x = as.factor(Variable), y = Value, fill = Value > 0)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "#D2B48C", "FALSE" = "#CD5C5C"),
                    labels = c("More men", "More women"))+
  labs(y = "Proportional Difference", title = "Proportional Difference by Variable") +
   theme(plot.background = element_rect(fill = "#983223"),
        panel.background = element_rect(fill = "#983223"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black"),
        axis.title.x = element_blank(),
        plot.title = element_text(color = "black"),
        legend.background = element_rect(color = "#983223"),
        legend.box.background = element_rect(color = "#983223"),
        text = element_text(size=15)) +
  theme(legend.background = element_rect(fill = "#983223", color = NA)) +
    theme(legend.text = element_text(color = "black"),        # Legend text color
          legend.title = element_text(color = "black")) +
  guides(fill = guide_legend(title = "Dominance"))


frmgham %>% 
  filter(DEATH==1) %>% 
  group_by(SEX) %>% 
  summarise("Stroke yes-no" = mean(STROKE),
            "Heart problems"=mean(MI_FCHD),
            "Diabetes"=mean(DIABETES),
            "Cigarettes"=mean(CIGPDAY, na.rm = TRUE),
            "Anti-hypertensive meds" = mean(BPMEDS, na.rm = TRUE)) -> pom  

# Compute proportional difference between Female and Male values for each variable
proportional_diff <- (pom[pom$SEX == "Female", -1] - pom[pom$SEX == "Male", -1]) / 
  (pom[pom$SEX == "Female", -1] + pom[pom$SEX == "Male", -1])

# Convert to a data frame and set column names
proportional_diff <- as.data.frame(proportional_diff)
colnames(proportional_diff) <- colnames(pom)[-1]

# Reshape the data into a long format for ggplot
proportional_diff_long <- proportional_diff %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value")

# Plot the proportional difference with color indicating sex
proportional_diff_long$Variable <- factor(proportional_diff_long$Variable,
                                          levels = c("Cigarettes", "Diabetes",
                                                     "Stroke yes-no", "Heart problems", 
                                                     "Anti-hypertensive meds"))


ggplot(proportional_diff_long, aes(x = as.factor(Variable), y = Value, fill = Value > 0)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "#D2B48C", "FALSE" = "#CD5C5C"),
                    labels = c("More men", "More women"))+
  labs(y = "Proportional Difference", title = "Proportional Difference by Variable") +
   theme(plot.background = element_rect(fill = "#983223"),
        panel.background = element_rect(fill = "#983223"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black"),
        axis.title.x = element_blank(),
        plot.title = element_text(color = "black"),
        legend.background = element_rect(color = "#983223"),
        legend.box.background = element_rect(color = "#983223"),
        text = element_text(size=15)) +
  theme(legend.background = element_rect(fill = "#983223", color = NA)) +
    theme(legend.text = element_text(color = "black"),        # Legend text color
          legend.title = element_text(color = "black")) +
  guides(fill = guide_legend(title = "Dominance"))


```


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)


differences <- data.frame(
  Hypertension = 15, # 15% higher risk of CVD for men with hypertension
  Alcohol = 10,     # 10% higher heart disease mortality rate among men who drink
  Obesity = -20,    # 20% higher risk of heart disease mortality among obese women
  Diabetes = -25,  # 25% higher risk of heart disease mortality among women with diabetes
  Smoking = -30  # 30% higher risk of heart disease mortality in smoking women than men
)


differences_long <- differences %>% 
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value")

differences_long$Variable <- factor(differences_long$Variable,
                                          levels = c(
                                                     "Smoking",
                                                     "Diabetes",
                                                     "Obesity",
                                                     "Alcohol",
                                            "Hypertension"
                                                     ))



ggplot(differences_long, aes(x = Value, y = as.factor(Variable), fill = Value<0)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "#D2B48C", "FALSE" = "#983223"),
                    labels = c("TRUE" = "Women", "FALSE" = "Men"))+
  labs(x="Difference in odds of dying due to heart disease between men and women", 
       title = "Ratio of dead of heart disease to all that were using/had:") +
  scale_x_continuous(labels = abs, limits = c(-40, 40), n.breaks = 11)+
   theme(
        plot.title = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 1, size=20),
        axis.text = element_text(color = "white"),
        axis.title = element_text(color = "white"),
        axis.title.x = element_text(color = "white", size=15),
        #axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(color = "white"),
        text = element_text(size=25),
        legend.title = element_blank(),
  panel.background = element_rect(fill = "#a09386",
                                  colour = NA_character_), # necessary to avoid drawing panel outline
  plot.background = element_rect(fill = "#a09386",
                                 colour = NA_character_), # necessary to avoid drawing plot outline
  legend.background = element_rect(fill = "#a09386", color = NA),
  legend.box.background = element_rect(fill = "#a09386", color = NA),
  legend.key = element_rect(fill = "#a09386")
)+ guides(fill = guide_legend(title = NA)) -> p
  
print(p)

  ggsave(
  plot = p,
  filename = "Men-woman-plot.png"#,bg = "transparent"
)

```

```{r}
frm2 <- frmgham

cosiek <- frm2 %>% filter(PERIOD == 1) %>%
  group_by(SEX) %>%
  summarise(cukier = sum(DIABETES == 1 & MI_FCHD == 1)/sum(DIABETES == 1),bez = sum(MI_FCHD)/n())
cosiek <- pivot_longer(cosiek,c(cukier,bez))
cosiek <- cosiek %>% mutate(SEX = case_when(SEX == 1 ~"mężczyzna",SEX == 2 ~"kobieta"))
cosiek$imie <- factor(paste(cosiek$name, cosiek$SEX), levels = c("bez kobieta","cukier kobieta", "bez mężczyzna", "cukier mężczyzna"))
ggplot(cosiek)+
  geom_bar(aes(x=imie, y = value), stat = "identity")
cosiek
```



```{r}
set.seed(1)
y <- rnorm(100)
df <- data.frame(
  x = 1,
  y0 = min(y),
  y25 = quantile(y, 0.40),
  y50 = mean(y),
  y75 = quantile(y, 0.60),
  y100 = max(y)
)
ggplot(df, aes(x)) +
  geom_boxplot(
   aes(ymin = y0, lower = y25, middle = y50, upper = y75, ymax = y100),
   stat = "identity"
 )
```

